---
title: "Daily Shelter and Overnight Services Availabilities"
#subtitle: "Marriage Data Cleaning"
author: "Cher Ning-Li"
thanks: "Code and data are available at: https://github.com/cher-ning/shelter_occupancies"
date: today
date-format: long
abstract: "In this paper I analyze the capacity vs real occupancy of shelters across Canada."
#TODO Complete abstract after paper.
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```


# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @citeTidyverse

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

Some of our data is of marriages (@fig-marriages), from @opendatatoronto.

```{r}
#| label: fig-occupancyrate-vs-month
#| fig-cap: Shelter Occupancy Rate in 2024 Over the Months
#| echo: false

#data |> 
#  ggplot(aes(x = date, y = occupancy_rate)) + 
#  geom_point()


hist(data$occupied_beds)
```

We can see that the number of marriages fluctuates across the years.


# Model


## Model set-up


We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false


```


# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details



\newpage


# References