---
title: "Week 3 Reflection"
subtitle: "Marriage Data Cleaning"
author: "Cher Ning-Li"
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "In this paper I analyze the number of marriages across the years, from 2018-2023."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```


# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

Some of our data is of marriages (@fig-marriages), from @OpenDataToronto.

```{r}
#| label: fig-marriages
#| fig-cap: Number of Marriages From 2018-2023
#| echo: false

data |> 
  ggplot(aes(x = date, y = marriage_licenses)) + 
  geom_point()
```

We can see that the number of marriages fluctuates across the years.


# Model


## Model set-up


We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```




# Discussion

## First discussion point {#sec-first-point}


## Second discussion point

## Third discussion point

## Weaknesses and next steps


\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check


## Diagnostics



\newpage


# References


